Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-09-24 formally.tex
pass through for the easy syntactic errors pointed out in the reviews. this closes #8.

Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions,
$\hlam{x}{\hexp}$. There is no type synthesis rule that applies to this
form, so lambda abstractions can appear only in analytic position,
i.e. where an expected type is known.\footnote{It is possible to also
  define a ``half-annotated'' synthetic lambda form, $\lambda x{:}\tau.e$,
  but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.}
Rule (\ref{rule:syn-lam}) is not quite the standard rule, reproduced below:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation*}
The problem is that this standard rule alone leaves us unable to analyze
lambda abstractions against the type hole, because the type hole is not
immediately of the form $\tarr{\htau_1}{\htau_2}$. There are two plausible
solutions to this problem. One solution would be to define a second rule
specifically for this case:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \tehole}{\hexp}{\tehole}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tehole}
}
\end{equation*}
Intead, we combine these two possible rules into a single rule through the
simple auxiliary \emph{matched arrow type} judgement,
$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure
\ref{fig:type-consistency}. This judgement leaves arrow types alone and
assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$. It
is easy to see that the two rules above are admissible by appeal to Rule
(\ref{rule:syn-lam}) and the matched arrow type judgement. Encouragingly,
this judgement also arises in the study of gradual type systems
\cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/GarciaC15,DBLP:conf/popl/RastogiCH12}.

>>>>

Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions,
$\hlam{x}{\hexp}$. There is no type synthesis rule that applies to this
form, so lambda abstractions can appear only in analytic position,
i.e. where an expected type is known.\footnote{It is possible to also
  define a ``half-annotated'' synthetic lambda form, $\lambda x{:}\tau.e$,
  but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.}
Rule (\ref{rule:syn-lam}) is not quite the standard rule, reproduced below:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation*}
The problem is that this standard rule alone leaves us unable to analyze
lambda abstractions against the type hole, because the type hole is not
immediately of the form $\tarr{\htau_1}{\htau_2}$. There are two plausible
solutions to this problem. One solution would be to define a second rule
specifically for this case:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \tehole}{\hexp}{\tehole}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tehole}
}
\end{equation*}
Instead, we combine these two possible rules into a single rule through the
simple auxiliary \emph{matched arrow type} judgement,
$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure
\ref{fig:type-consistency}. This judgement leaves arrow types alone and
assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$. It
is easy to see that the two rules above are admissible by appeal to Rule
(\ref{rule:syn-lam}) and the matched arrow type judgement. Encouragingly,
this judgement also arises in the study of gradual type systems
\cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/GarciaC15,DBLP:conf/popl/RastogiCH12}.

### 2016-09-24 hazelnut-popl17.tex
working on #6
pass through for the easy syntactic errors pointed out in the reviews. this closes #8.

\def\OPTIONConf{1}%
%\newif\iftr\trfalse
\newif\iftr\trtrue
%\documentclass[12pt]{article}
\RequirePackage{etex}
\documentclass[preprint,9pt]{sigplanconf}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{multicol}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.9}
\usepackage{soul}
\setulcolor{red}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\usepackage{listings}
\lstset{tabsize=2,
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont}
\usepackage{microtype}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

The examples given so far either define a language with a trivial static
semantics, or do not attempt to maintain well-typedness as an edit
invariant. This can pose problems, for reasons discussed in the
Introduction. One apparent exception is Unison~\cite{unison}, a structure
editor for a typed functional language similar to Haskell. Like Hazelnut,
it seems to define some notion of well-typedness for expressions with holes
(though there is no technical documentation on virtually any aspect of its
design.) Unlike Hazelnut, it does not have a notion analagous to Hazelnut's
notion of a non-empty hole. As such, programmers must construct programs in
a rigid outside-in manner, as discussed in Sec. \ref{sec:example}. Another
system with the same problem is Polymorphic Blocks, a block-based user
interface where the structure of block connectors encodes a
type \cite{DBLP:conf/chi/LernerFG15}.

@@

\subsection{Gradual Type Systems}
A significant contribution of this paper is the discovery of a clear
technical relationship between typed structure editing and gradual
typing. In particular, the machinery necessary to give a reasonable
semantics to type holes -- i.e. type consistency and type matching --
coincides with that developed in the study of gradual type systems for
functional languages. The pioneering work of Siek and Taha \cite{Siek06a}
introduced type consistency. Subsequent work developed the concept of type
matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15}. In
retrospect, this relationship is perhaps unsurprising: gradual typing is,
notionally, also motivated by the idea of iterated development of a program
where every intermediate state is well-defined in some sense.
%Typed structure editing endows this intuitive notion with technical force.

Recent work has discovered a systematic procedure for generating a
``gradual version'' of a standard type
system \cite{DBLP:conf/popl/CiminiS16}. This system, called the
Gradualizer, operates on a logic program that specifies a simple type
assignment system with some additional annotations to generate a
corresponding specification of a gradual type system. The authors leave
support for working with bidirectional type systems as future work. This
suggests the possibility of an analagous ``Editorializer'' that generates a
specification of a typed structure editor from a simple language
definition. Our exposition in Sec. \ref{sec:extending} certainly suggests
that many of the necessary definitions follow seemingly mechanically from
the definition of the static semantics, and the relationship with gradual
typing suggests that many of the technical details of this transformation
may already exist in the Gradualizer. One possibility we have explored
informally is to use Agda's reflection features to implement such a system.

@@

% \todo{Eclipse parsing partial programs // other heuristics}

% \paragraph{Gradual typing: Has Matched Relation}
% %
% A. Rastogi, A. Chaudhuri, and B. Hosmer. The ins and outs of gradual type inference.
% \\
% Definition 3.1 is similar to our ``has matched'' relations:
% \\
% \url{https://www.cs.umd.edu/~avik/papers/iogti.pdf}

% This paper defines ``has matched'' relation for arrow, just like us:
% \\
% \url{http://cimini.info/publications/Gradualizer_Draft.pdf}
% \\
% They credit this technique to the following paper:
% \\
% \url{http://www.cs.ubc.ca/~rxg/ptsgp.pdf}
% \\
% But, the Rastogi paper came earlier.

% \todo{TODO: \url{http://research.nii.ac.jp/~hu/pub/hosc07.pdf}}

% Agda and Idris are two dependently typed languages that attempt to simulate a structured editor from within a rich text editor (e.g. Emacs.) These systems also have notions of holes and use types to guide the user toward filling these holes. These  systems are also, to our knowledge, not formally well-defined but rather exist only as part of system implementations.

% Our work differs from all of these in that we begin with a formal editor calculus and build from there, rather than starting with an implementation and leaving many of the formal details formally unspecified. For example, while Lamdu has many interesting features, there is no theoretical basis presented for their work -- it is a rather large body of Haskell code with an unclear (and indeed, often somewhat perplexing, in our experience) action model. Unison is also a rather large body of Haskell code, though its action model appears superficially more similar to ours. We maintain what we believe to be a stronger action sensibility invariant than Unison (i.e. in Unison, one must construct expressions from the outside-in.) These systems are rich sources of interesting ideas, however -- there is room enough for many different approaches in this (re-)emerging space.

% \todo{TODO: cite \url{http://cseweb.ucsd.edu/~lerner/pb.html}}

% \todo{TODO: cite} \url{http://delivery.acm.org/10.1145/1060000/1056965/p1557-ko.pdf?ip=73.154.143.34&id=1056965&acc=AUTHOR-IZED&key=4D4702B0C3E38B35%2E4D4702B0C3E38B35%2E4D4702B0C3E38B35%2EC2CC0A83071F7605&CFID=780613712&CFTOKEN=33786941&__acm__=1462392463_f29fc98965004c61bfb1291d07756a23}

%Drag-and-drop / for novices: lots of examples, e.g. Alice and others
%
%Contemporary: Lamdu, MPS/Mbeddr, TouchDevelop
%
%Hybrid: Cyrus' active code completion paper

%\subsection{Refactoring Models}
%(Michael, can you fill this section out?)

%\subsection{Formal Editor Models}
%Need to do a search to see what else has been done...

@@

% By keeping the program in both a structurally and semantically
% well-defined state at all times, Hazelnut allows users to avoid premature
% commitment~\cite{green1996usability}.  By inserting a hole, the user can
% leave certain parts of the program unfinished, \todo{What is the best
% word to define this concept?  In progress, unfinished, incomple?} and yet
% still in a well defined type state.  This also enables progressive
% evaluation~\cite{green1996usability}, because unfinished solutions are
% also are well-defined at all times, thus enabling tools to provide
% evaluation of unfinished solutions.\todo{integrate this}

@@

% We already discussed a connection to gradual typing \cite{Siek06a}. We
% hope to explore this connection more thoroughly. In particular, it may be
% possible to better support exploratory and live programming by allowing
% even programs with holes in them to execute as long as those holes are
% only in the type portions, by deferring to the semantics given in work on
% gradual typing.

% It may also be possible to give a dynamics to incomplete
% expressions. Prior work on staged evaluation suggests that there may be a
% connection to modal logic, viewing holes as quantifying over all possible
% terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}. In
% developing a dynamic semantics, we will also need to handle terms like
% $\hhole{\hehole}$ and $\hhole{\hhole{\hexp}}$. In our semantics given
% here, we eliminated them as they came up in a somewhat \emph{ad hoc}
% manner. We have not yet explored an equational theory for terms with
% holes, but intend to once our formalization effort is more
% mature.\todo{fix bib}

% \todo{new action form that makes actions extensible given proofs of
% admissiblity of a derived form, like prevSib}

>>>>

\def\OPTIONConf{1}%
\newif\iftr\trtrue
\RequirePackage{etex}
\documentclass[preprint,9pt]{sigplanconf}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{multicol}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.9}
\usepackage{soul}
\setulcolor{red}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\usepackage{listings}
\lstset{tabsize=2,
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont}
\usepackage{microtype}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

The examples given so far either define a language with a trivial static
semantics, or do not attempt to maintain well-typedness as an edit
invariant. This can pose problems, for reasons discussed in the
Introduction. One apparent exception is Unison~\cite{unison}, a structure
editor for a typed functional language similar to Haskell. Like Hazelnut,
it seems to define some notion of well-typedness for expressions with holes
(though there is no technical documentation on virtually any aspect of its
design.) Unlike Hazelnut, it does not have a notion analogous to Hazelnut's
notion of a non-empty hole. As such, programmers must construct programs in
a rigid outside-in manner, as discussed in Sec. \ref{sec:example}. Another
system with the same problem is Polymorphic Blocks, a block-based user
interface where the structure of block connectors encodes a
type \cite{DBLP:conf/chi/LernerFG15}.

@@

\subsection{Gradual Type Systems}
A significant contribution of this paper is the discovery of a clear
technical relationship between typed structure editing and gradual
typing. In particular, the machinery necessary to give a reasonable
semantics to type holes -- i.e. type consistency and type matching --
coincides with that developed in the study of gradual type systems for
functional languages. The pioneering work of Siek and Taha \cite{Siek06a}
introduced type consistency. Subsequent work developed the concept of type
matching \cite{DBLP:conf/popl/RastogiCH12,DBLP:conf/popl/GarciaC15}. In
retrospect, this relationship is perhaps unsurprising: gradual typing is,
notionally, also motivated by the idea of iterated development of a program
where every intermediate state is well-defined in some sense.

Recent work has discovered a systematic procedure for generating a
``gradual version'' of a standard type
system \cite{DBLP:conf/popl/CiminiS16}. This system, called the
Gradualizer, operates on a logic program that specifies a simple type
assignment system with some additional annotations to generate a
corresponding specification of a gradual type system. The authors leave
support for working with bidirectional type systems as future work. This
suggests the possibility of an analogous ``Editorializer'' that generates a
specification of a typed structure editor from a simple language
definition. Our exposition in Sec. \ref{sec:extending} certainly suggests
that many of the necessary definitions follow seemingly mechanically from
the definition of the static semantics, and the relationship with gradual
typing suggests that many of the technical details of this transformation
may already exist in the Gradualizer. One possibility we have explored
informally is to use Agda's reflection features to implement such a system.

@@

@@

@@
